---
title: "Evaluate effectiveness of Intelligent Sensor System - EDA report"
author: "Xing Song"
date: "2/10/2021"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
```


```{r load_lib}
source("../R/util.R")
#install specific packages
require_libraries(c("tidyverse",
                    "magrittr",
                    "broom",
                    "Hmisc",
                    "lme4",
                    "lmerTest",
                    "kableExtra",
                    "ggstance"
                    ),
                  verb=FALSE)
```


```{r load_data}
enroll<-readRDS("../data/preproc/enroll.rda") 
walkspeed<-readRDS("../data/preproc/walkspeed.rda")
gaitrite<-readRDS("../data/preproc/gaitrite.rda")
grip<-readRDS("../data/preproc/grip.rda")
```


```{r overview}
pat_n<-length(unique(enroll$Research_ID))
case_n<-length(unique(enroll$Research_ID[enroll$Summary_Type=="Linguistic"]))
ctrl_n<-length(unique(enroll$Research_ID[enroll$Summary_Type=="Standard"]))
```

Among all `r pat_n` enrolled subjects, `r case_n` are randomized to the "Linguistic" (shortened as "L") group and `r ctrl_n` were randomized to the "Standard" (shortened as "S") group. 


### Table 1 - Basic Characteristic Comparison

```{r tbl1}
enroll %<>%
  left_join(walkspeed %>% group_by(Research_ID) %>%
              arrange(Date) %>% slice(1:1) %>%
              select(Research_ID,ws_10ft),
            by="Research_ID") %>%
  left_join(grip %>% group_by(Research_ID) %>%
              arrange(Date) %>% slice(1:1) %>%
              select(Research_ID,grip_right,grip_left),
            by="Research_ID") %>%
  left_join(gaitrite %>% group_by(Research_ID,var) %>%
              arrange(Date_Time_of_Test) %>% slice(1:1) %>%
              select(Research_ID,var,val) %>%
              spread(var,val),
            by="Research_ID")

var_tbl1<-tibble(var_nm=as.character(),
                 var_type=as.character())
var_tbl1 %<>% 
  add_row(var_nm="Age_at_Enr",var_type="num") %>%
  add_row(var_nm="Sex",var_type="cat") %>%
  add_row(var_nm="HT",var_type="num") %>%
  add_row(var_nm="Wheelchair",var_type="cat") %>%
  add_row(var_nm="System_Installed",var_type="cat") %>%
  add_row(var_nm="Censored",var_type="num") %>%
  add_row(var_nm="Enroll_Days",var_type="num") %>%
  add_row(var_nm="Site",var_type="cat") %>%
  bind_rows(data.frame(var_nm=c("ws_10ft","grip_right","grip_left",unique(gaitrite$var)),
                       var_type=rep("num",85),
                       stringsAsFactors = F))

tbl1<-univar_analysis_mixed(id=enroll$Research_ID,
                            grp=enroll$Summary_Type,
                            X=enroll[,var_tbl1$var_nm],
                            data_type=var_tbl1$var_type,
                            pretty=T)

tbl1 %>%
  kbl(caption="Table 1 - Demographic and Clinical Characteristic Comparison") %>%
  collapse_rows(columns = c(1,5), valign = "top") %>%
  kable_material(c("striped", "hover"))
```

No significant differences were observed between the L and S group.  


***************************************************************************************************************


### Walking Speed (10ft)

```{r ws_overview}
ws_pat_n<-length(unique(walkspeed$Research_ID))
ws2_pat_n<-length(unique(walkspeed$Research_ID[walkspeed$Repeat_Meas_n>1]))
ws2_case_n<-length(unique(walkspeed$Research_ID[walkspeed$Repeat_Meas_n>1&walkspeed$Summary_Type=="Linguistic"]))
ws2_ctrl_n<-length(unique(walkspeed$Research_ID[walkspeed$Repeat_Meas_n>1&walkspeed$Summary_Type=="Standard"]))
```

The 10-ft walking speed is measured every 6 months, 2 measurement were taken each time, and the average value is used for final analysis. `r ws_pat_n` subjects has at least some walking speed records, while only `r ws2_pat_n` had more than 1 repeated measures, `r ws2_case_n` from L group and `r ws2_ctrl_n` from S group.


```{r ws_exp1}
ggplot(walkspeed %>%
         filter(Repeat_Meas_n > 1) %>%
         select(Research_ID,ws_10ft,Mth3_Since_Last,Summary_Type),
       aes(x=Mth3_Since_Last,y=ws_10ft,color=Summary_Type)) +
  geom_point(alpha=0.5) +
  geom_smooth(aes(group=Research_ID),
              method="lm",formula = 'y~x',se=FALSE,linetype=1)

```

```{r ws_exp2}
#collect personalized rates of change
walkspeed_summ<-walkspeed %>%
  filter(Repeat_Meas_n > 1) %>%
  select(Research_ID,ws_10ft,Mth_Since_Last) %>%
  nest(data=-Research_ID) %>%
  mutate(
    ws_lm=map(data,~lm(ws_10ft~Mth_Since_Last,data=.x)),
    ws_lm_tidied=map(ws_lm,tidy)
  ) %>%
  unnest(ws_lm_tidied) %>%
  select(-data,-ws_lm) %>%
  filter(term=="Mth_Since_Last") %>%
  select(Research_ID,estimate,p.value) %>%
  mutate(p.value=case_when(p.value==NaN ~ 0,
                           TRUE ~ p.value)) %>%
  mutate(change_dir=case_when(p.value > 0.05 ~ 'No Change',
                              estimate/abs(estimate)< 0 ~ 'Decrease',
                              TRUE ~ 'Increase'),
         estimate=round(estimate,3)) %>%
  left_join(enroll %>%
              select(Research_ID,Summary_Type,Site),
            by="Research_ID")


ggplot(walkspeed_summ, aes(x = estimate, y = 1)) +
  # horizontal box plot
  geom_boxploth(aes(fill = Summary_Type)) +
  # normal density plot
  geom_density(aes(x = estimate,y=..scaled..), inherit.aes = FALSE) +
  # vertical lines at Q1 / Q2 / Q3
  stat_boxploth(geom = "vline", aes(xintercept = ..xlower..),linetype=2) +
  stat_boxploth(geom = "vline", aes(xintercept = ..xmiddle..),linetype=2) +
  stat_boxploth(geom = "vline", aes(xintercept = ..xupper..),linetype=2) +
  # annotate boxplots
  stat_summaryh(geom="text", fun.x=quantile,
                aes(label=sprintf("%1.1f", ..x..)),
                position=position_nudge(y=0.2),
                size=3.5,angle=90) +
  
  facet_grid(Summary_Type ~ ., scales="free") +
  labs(x="Rate of Walking Speed change per Month", 
       y="density")
```


```{r ws_exp2_site,fig.width=10}

ggplot(walkspeed_summ, aes(x = estimate, y = 1)) +
  # horizontal box plot
  geom_boxploth(aes(fill = Summary_Type, linetype = Site)) +
  # normal density plot
  geom_density(aes(x = estimate,y=..scaled..), inherit.aes = FALSE) +
  # vertical lines at Q1 / Q2 / Q3
  stat_boxploth(geom = "vline", aes(xintercept = ..xlower..),linetype=2) +
  stat_boxploth(geom = "vline", aes(xintercept = ..xmiddle..),linetype=2) +
  stat_boxploth(geom = "vline", aes(xintercept = ..xupper..),linetype=2) +
  # annotate boxplots
  stat_summaryh(geom="text", fun.x=quantile,
                aes(label=sprintf("%1.1f", ..x..)),
                position=position_nudge(y=0.2),
                size=3.5,angle=90) +
  
  facet_grid(Summary_Type + Site ~ ., scales="free") +
  labs(x="Rate of Walking Speed change per Month", 
       y="density")
```

Based on the estimated personal rate of changes, we can group the patients into the following three categories: 
1. "No Change": when estimated slope is _not statistically significantly different_ from 0 (p.value > 0.05)
2. "Increase": when estimated slope is _statistically significantly greater than_ 0 (p.value <= 0.05)
3. "Decrease": when estimated slope is _statistically significantly less than_ 0 (p.value <= 0.05)

```{r}
tbl2<-univar_analysis_mixed(id=walkspeed_summ$Research_ID,
                            grp=walkspeed_summ$Summary_Type,
                            X=walkspeed_summ[,c("estimate","change_dir")],
                            data_type=c("num","cat"),
                            pretty=T,
                            digit=3)

tbl2 %>%
  kbl(caption="Table 2 - Walking Speed Change per Month") %>%
  collapse_rows(columns = c(1,5), valign = "top") %>%
  kable_material(c("striped","hover"))
```

From the boxplots and summary table above, we observe:
- There are no significant difference of Walking Speed change over time between two groups, however, the walking speed change is numerically more stable among the "Linguistic" group;
- The walking speed change is a lot more variable among the the "Standard" group;due to such variability difference, we observe more extremely worse cases in the control ("Standard") group


We further build a mixed effect model to better account for the nested variability structure, but still observed no significant difference in walking speed change between the two comparison groups.

```{r ws_lmm}
fit_ws<-lmer(ws_10ft ~ Mth3_Since_Last*Summary_Type+(1+Mth3_Since_Last|Research_ID),
             data=walkspeed %>% filter(Repeat_Meas_n > 1))
summary(fit_ws)
confint(fit_ws)
anova(fit_ws)
```



***************************************************************************************************************


### Grip Strength

```{r grip_overview}
grip_pat_n<-length(unique(grip$Research_ID))
grip2_pat_n<-length(unique(grip$Research_ID[grip$Repeat_Meas_n>1]))
grip2_case_n<-length(unique(grip$Research_ID[grip$Repeat_Meas_n>1&grip$Summary_Type=="Linguistic"]))
grip2_ctrl_n<-length(unique(grip$Research_ID[grip$Repeat_Meas_n>1&grip$Summary_Type=="Standard"]))
```

The grip strength was measured every 3 months, 2 measurements were taken each time, and the average value is used for final analysis. `r grip_pat_n` subjects has at least some walking speed records, while only `r grip2_pat_n` had more than 2 repeated measures over time, `r grip2_case_n` from L group and `r grip2_ctrl_n` from S group.

```{r grip_plot}
ggplot(grip %>%
         filter(Repeat_Meas_n > 1) %>%
         select(Research_ID,grip_right,grip_left,Mth3_Since_Last) %>%
         inner_join(enroll %>% select(Research_ID,Summary_Type),
                    by="Research_ID") %>%
         gather(grip_side,grip,-Research_ID,-Summary_Type,-Mth3_Since_Last),
       aes(x=Mth3_Since_Last,y=grip,color=Summary_Type)) +
  geom_point(alpha=0.5) +
  geom_smooth(aes(group=Research_ID),method="lm",
              formula = 'y~x',se=FALSE,linetype=1) +
  facet_wrap(~grip_side,ncol=2,scales="free")
```


```{r grip_grp}
#collect personalized rates of change
grip_summ<-grip %>%
  filter(Repeat_Meas_n > 1) %>%
  select(Research_ID,grip_right,grip_left,Mth_Since_Last) %>%
  gather(grip_side,grip,-Research_ID,-Mth_Since_Last) %>%
  filter(!is.na(grip)) %>% #not always observed for both hands
  nest(data=c(grip,Mth_Since_Last)) %>%
  mutate(
    grip_lm=map(data, ~lm(grip ~ Mth_Since_Last,data=.x)),
    grip_lm_tidied=map(grip_lm,tidy)
  ) %>%
  unnest(grip_lm_tidied) %>%
  select(-data,-grip_lm) %>%
  filter(term=="Mth_Since_Last") %>%
  select(Research_ID,grip_side,estimate,p.value) %>%
  mutate(p.value=case_when(p.value==NaN ~ 0,
                           TRUE ~ p.value)) %>%
  mutate(change_dir=case_when(p.value > 0.05 ~ 'No Change',
                              estimate/abs(estimate)< 0 ~ 'Decrease',
                              TRUE ~ 'Increase'),
         estimate=round(estimate,3)) %>%
  left_join(enroll %>%
              select(Research_ID,Summary_Type),
            by="Research_ID")


ggplot(grip_summ, aes(x = estimate, y = 1)) +
  # horizontal box plot
  geom_boxploth(aes(fill = Summary_Type)) +
  # normal density plot
  geom_density(aes(x = estimate,y=..scaled..), inherit.aes = FALSE) +
  # vertical lines at Q1 / Q2 / Q3
  stat_boxploth(geom = "vline", aes(xintercept = ..xlower..),linetype=2) +
  stat_boxploth(geom = "vline", aes(xintercept = ..xmiddle..),linetype=2) +
  stat_boxploth(geom = "vline", aes(xintercept = ..xupper..),linetype=2) +
  # annotate boxplots
  stat_summaryh(geom="text", fun.x=quantile,
                aes(label=sprintf("%1.1f", ..x..)),
                position=position_nudge(y=0.2),
                size=3.5,angle=90) +
  facet_grid(grip_side + Summary_Type ~ ., scales="free") +
  labs(x="Grip Strength change per Month", 
       y="density")
```

**Right Grip**

```{r}
grip_summ_r<-grip_summ %>% filter(grip_side=="grip_right")
tbl3<-univar_analysis_mixed(id=grip_summ_r$Research_ID,
                            grp=grip_summ_r$Summary_Type,
                            X=grip_summ_r[,c("estimate","change_dir")],
                            data_type=c("num","cat"),
                            pretty=T,
                            digit=3)

tbl3 %>%
  kbl(caption="Table 3 - Right Grip Change per Month") %>%
  collapse_rows(columns = c(1,5), valign = "top") %>%
  kable_material(c("striped","hover"))

```

```{r gripr_lmm, eval=F}
fit_gripr<-lmer(grip_right ~ Mth_Since_Last*Summary_Type+(1|Research_ID),
                data=grip %>% filter(Repeat_Meas_n > 1))
summary(fit_gripr)
anova(fit_gripr)
confint(fit_gripr)
```


**Left Grip**

```{r gripl_lmm, eval=F}
fit_gripl<-lmer(grip_left ~ Mth3_Since_Last*Summary_Type+(1|Research_ID),
                data=grip %>% filter(Repeat_Meas_n > 1))
summary(fit_gripl)
anova(fit_gripl)
confint(fit_gripl)
```


```{r}
grip_summ_l<-grip_summ %>% filter(grip_side=="grip_left")
tbl4<-univar_analysis_mixed(id=grip_summ_l$Research_ID,
                            grp=grip_summ_l$Summary_Type,
                            X=grip_summ_l[,c("estimate","change_dir")],
                            data_type=c("num","cat"),
                            pretty=T,
                            digit=3)

tbl4 %>%
  kbl(caption="Table 4 - Left Grip Change per Month") %>%
  collapse_rows(columns = c(1,5), valign = "top") %>%
  kable_material(c("striped","hover"))

```

The "No change", "Increase", and "Decrease" groups are defined the similar way as above. By comparing the personalized rate of change for grip strength, we observed that **there right grip strength reduced with a rate that is significantly slower than the standard group**. 



***************************************************************************************************************


### GaitRite Measurement

```{r gait_overview}
gaitrite %<>%
  mutate(var_idx=dense_rank(var))

gait_p<-max(gaitrite$var_idx)
gait_pat_n<-length(unique(gaitrite$Research_ID))
gait2_pat_n<-length(unique(gaitrite$Research_ID[gaitrite$Repeat_Meas_n>1]))
gait2_case_n<-length(unique(gaitrite$Research_ID[gaitrite$Repeat_Meas_n>1&gaitrite$Summary_Type=="Linguistic"]))
gait2_ctrl_n<-length(unique(gaitrite$Research_ID[gaitrite$Repeat_Meas_n>1&gaitrite$Summary_Type=="Standard"]))
```

GaitRite metrics were collected every 6 months with `r gait_p` different metrics reported. `r gait_pat_n` subjects has at least some walking speed records, while only `r gait2_pat_n` had more than 2 repeated measures over time, `r gait2_case_n` from L group and `r gait2_ctrl_n` from S group.


```{r gait_discovery}
test_df<-c()
sig_plot<-list()
for(i in seq_len(gait_p)){
  #get a cut of the data set 
  data_i<-gaitrite %>% filter(Repeat_Meas_n > 1 & var_idx == i) %>% unique
  var_i<-data_i %>% select(var) %>% slice(1:1) %>% unlist
  
  #build glmm model with random intercept and slope
  fit_gait<-lmer(val ~ Mth3_Since_Last*Summary_Type+(1+Mth3_Since_Last|Research_ID),
                 data=data_i)
  
  #get model summary
  summ<-summary(fit_gait)
  # ci<-confint(fit_gait)
  
  test_df_add<-cbind(
    var_i
    ,round(summ$coefficients[2:4,],4)
    # ,ci[6:8,]
  )
  
  #collect visual for significant-variant factors
  if(min(summ$coefficients[3:4,5])<=0.1){
    test_df_add<-cbind(test_df_add,sig=1)
    
    sig_plot[[var_i]]<-ggplot(data_i,
                              aes(x=Day_Since_Last,y=val,color=Summary_Type)) +
      geom_point(alpha=0.5) +
      geom_smooth(aes(group=Research_ID),
                  method="lm",formula = 'y~x',se=FALSE,
                  linetype=1) +
      facet_wrap(~var,ncol=4,scales="free")
  }else{
    test_df_add<-cbind(test_df_add,sig=0)
  }
  
  test_df<-rbind(test_df,test_df_add)
}
#re-organize the table for better reasability
rownm<-rownames(test_df)
row.names(test_df)<-NULL
test_df2show<-data.frame(
  GaitRite_meas=test_df[,1],
  coef=rownm,
  test_df[,-1]) %>%
  group_by(GaitRite_meas) %>%
  dplyr::mutate(min_pval=min(`Pr...t..`)) %>%
  ungroup %>%
  arrange(min_pval,GaitRite_meas) %>%
  select(-min_pval)
  
test_df2show %>%
  filter(sig==1) %>%
  kbl(caption="Table 2 - GLMM-Model-based screening results of GaitRite metrics") %>%
  collapse_rows(columns = 1, valign = "top")

gait_p_sel<-length(sig_plot)
```

I performed mixed-model-based screening over all the `r gait_p` gaitrite variables, and presented the results in Table 2. `r gait_p_sel` has showed some significance (p-value <= 0.1) in either overall difference of metrics or change of the metrics over time. 

```{r gait_plot}
sig_nm<-names(sig_plot)
for(nm in sig_nm){
  print(sig_plot[[nm]])
}
```